---
title: "Introduction to R"
author: 
- "Markus List (<markus.list@wzw.tum.de>)</br></br>Head of the Big Data in BioMedicine Group</br>Chair of Experimental Bioinformatics</br>Technical University of Munich</br></br><https://biomedical-big-data.de></br>"
date: "05.12.2019"
output: ioslides_presentation
logo: ../tum.png
css: ../biomedbigdata.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Shiny

- turn R scripts into interactive web applications
- make package functionality available to non-programmers
- visualize data / results
- filter data / results
- much easier than a traditional web framework
- many examples, great community
- issue: no multi-user support

## Structure

Shiny apps are contained in a single script called app.R. 

app.R has three components:

 - a user interface object
 - a server function
 - a call to the shinyApp function

## ui

```{r, eval = FALSE, echo = TRUE}
library(shiny)
# Define UI for app that draws a histogram ----
ui <- fluidPage(
  # App title ----
  titlePanel("Hello Shiny!"),
  # Sidebar layout with input and output definitions ----
  sidebarLayout(
    # Sidebar panel for inputs ----
    sidebarPanel(
      # Input: Slider for the number of bins ----
      sliderInput(inputId = "bins", label = "#bins:", min = 1, max = 50, value = 30)),
    # Main panel for displaying outputs ----
    mainPanel(
      # Output: Histogram ----
      plotOutput(outputId = "distPlot")
    )))
```

## server

```{r, echo = TRUE, eval = FALSE}
server <- function(input, output) {
  output$distPlot <- renderPlot({

    x    <- faithful$waiting
    bins <- seq(min(x), max(x), length.out = input$bins + 1)

    hist(x, breaks = bins, col = "#75AADB", border = "white",
         xlab = "Waiting time to next eruption (in mins)",
         main = "Histogram of waiting times")
    })}
```

## run the app

```{r, echo = TRUE, eval = FALSE}
library(shiny)
ui <- ...

server <- ...

shinyApp(ui = ui, server = server)
```

or 

```{r, echo = TRUE, eval = FALSE}
library(shiny)
runApp("directory")
```

## Examples

Shiny has 11 built in examples

```{r, echo = TRUE, eval = FALSE}
library(shiny)
runExample("01_hello")
```


## Challenge

Build a shiny app that uses your package to

- load a tabula muris FACS file
- visualizes the fist two principle components
- think of other useful plots

Alternative: be creative!

Bonus tasks:

- uses the top 500 most variable genes and plot a heatmap
- show a table that allows you to filter expression values
- use cell type annotation to color the cells