---
title: "R packages"
author: 
- "Markus List (<markus.list@wzw.tum.de>)</br></br>Head of the Big Data in BioMedicine Group</br>Chair of Experimental Bioinformatics</br>Technical University of Munich</br></br><https://biomedical-big-data.de></br>"
date: "05.12.2019"
output: ioslides_presentation
logo: ../tum.png
css: ../biomedbigdata.css
---

```{r setup, include=FALSE}
```

## Introduction 

  - In R, the fundamental unit of shareable code is the package.
  - A package bundles together code, data, documentation, and tests
  - Packages are easy to share with others. 
  - This huge variety of packages is one of the reasons that R is so successful

  - Install from CRAN with `install.packages("x")`.
  - Use `library("x")`.
  - Help `package?x` and `help(package = "x")`.

## Why

  - Share code in a usable format
  - Save time with conventions

## Philosophy 

  - Anything that can be automated, should be automated. 
  - Devtools package to simplify development

## Package states

When you create or modify a package, you work on its "source code" or "source files". This is NOT the package form you are most familiar with from day-to-day usage. Package development workflows make much more sense if you understand the five states an R package can be in:

  * source
  * bundled
  * binary
  * installed
  * in-memory
  
## Package libraries 

In R, a __library__ is a directory containing installed packages,

> We use the `library()` function to load a __package__. 

The distinction between the two is important and useful as you get involved in package development.

You can have multiple libraries on your computer. 

```{r, eval = TRUE}
.libPaths()
unlist(lapply(lapply(.libPaths(), list.dirs, recursive = FALSE, full.names = FALSE), length))
```

## Prepare your system 

```{r, eval = FALSE}
install.packages(c("devtools", "roxygen2", "testthat", "knitr"))
```

> Do this now please!

## Loading source code from a package
Example of how to simulate installing and loading a package, during interactive development:

```{r eval = FALSE}
library(devtools)
load_all()
```

### Personal startup configuration

```{r, eval = FALSE}
library(devtools)
```

## R build toolchain

To be fully capable of building R packages from source that contain C, C++ or Fortran code, you'll also need a compiler and a few other command line tools. 

- Windows: install Rtools from <https://cran.r-project.org/bin/windows/Rtools/> and run the installer.
- macOS: install the Xcode command line tools, which requires that you [register as an Apple developer](https://developer.apple.com/programs/register/) 

Then, in the shell, do:

```shell
xcode-select --install
```

- linux: install R development tools. For example, on Ubuntu (and Debian) you need to install the `r-base-dev` package.

## Verify system prep

You can check that you have everything installed and working by running the following code:

```{r, eval = TRUE}
library(devtools)
has_devel()
```

## Package structure

- Code (R/)
- Package metadata (DESCRIPTION)
- Object documentation (man/)
- Vignettes (vignettes/)
- Testing (tests/)
- Namespaces (NAMESPACE)
- Data (data/)
- Compiled code (src/)
- Installed files (inst/)

## Making code available

- version control system
- example: git, gitlab, github
- shinyapps.io

## Testing

- good code comes with unit tests
- rule of thumb is to have more tests than actual code
- test whatever you can think of!
- testthat R package supports you in proper testing

```{r}
usethis::use_test("name")
```


## More literature

<http://r-pkgs.had.co.nz/>

## Live Example

Fingers crossed

## Challenge:

Create an R package with the following functions

```{r, echo = TRUE, eval = FALSE}
read_tabula_muris <- function(file){ ... }

plot_pca <- function(counts, pc_1, pc_2){ ... }

convert_to_sce <- function(counts) { #return a SingleCellExperiment}

convert_cell_ids <- function(counts) { #extract meta data from cell ids}
```
